Safe Transferable Regions

### 2018-03-20 paper/type-system.tex
Ecoop18-related changes and the rebuttal.

letregion $\pi_0$ in
  let snd = new Object<$\pi_0$>() in
  letregion $\pi_1$ in
    let fst = new Object<$\pi_1$>() in
    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
and $\pi_1$ are live during the instantiation, and (b). $\pi_0
\outlives \pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\pi_1$, and (b). it
holds references to objects allocated in region $\pi_0$ and $\pi_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\pi_1$, then \C{p} would be contained in $\pi_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
$B\inang{\tbar}@\pi$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\pi$. Class fields are not allowed to
be type polymorphic, in keeping with C\#, which constitutes the core
of \name.

Note that classes in \FB are independently parameterized over types
and regions. While this design decision has a downside that it allows
type variables to only denote the types of objects contained in a
single region, it yields benefits that far outweigh the costs. In
particular, it lets us support region-polymorphic higher-order
functions as class fields.  This allows, for example, a generic class,
whose type parameters are $a$ and $b$, to contain a region-polymorphic
function of type
$\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0 \conj
\rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
to the point that inference becomes practical
(Sec.~\ref{sec:type-inference}); as our experience with Hindley-Milner
type systems suggests, type inference is pivotal to the practical
utility of sophisticated type systems.

@@

Well-formedness and typing rules of \fbname establish the conditions
under which a region type is considered well-formed, and an expression
is considered to have a certain region type, respectively.
Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
rules\footnote{Full formal development can be found in the
appendix}. The rules refer to a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of regions that are estimated to be
live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their \emph{bounds}, i.e., classes they are declared to
extend (this artifact is inherited from FGJ), and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
In addition, the context for the expression typing judgment also
includes (a). a type environment ($\env \in x \mapsto \tau$) that
contains the type bindings for variables in scope, and (b). The region
($\rgn$) that serves as the allocation context for the expression
being type checked. Like the judgments in FGJ~\cite{fgj}, all the
judgments defined by the rules in Fig.~\ref{fig:fb-staticsem} are
implicitly parameterized on a class table ($CT \in cn \mapsto D$) that
maps class names to their definitions in \FB.

@@

\noindent The well-formedness judgment on region types
($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
judgments on core types. We use a double-piped turnstile ($\Vdash$)
for judgments in FGJ~\cite{fgj}, and a simple turnstile ($\vdash$) for
those in \FB. The class table ($\absof{CT}$) for FGJ judgments is
derived from \FB's class table ($CT$) by erasing all region
annotations on types, and region arguments in expressions
($\absof{\cdot}$ denotes the region erasure operation). 
% For judgments in FGJ, we use a
% class table that maps class names to their region-erased definitions.
% We denote the region-erased class table as ($\absof{CT}$). The precise
% semantics of region erasure ($\absof{\cdot}$) are defined
% in~\cite{techrep}, but it can be understood as an operation that
% erases region annotations on types. The class table is assumed to bind
% the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
% This mostly eliminates the need to treat $\RgnZ$ objects, which double
% up as transferable region handlers, as a special case in semantics.
The well-formedness rule for class types
($B\inang{\tbar}\inang{\rbar}$) is responsible for enforcing the
safety property that prevents objects from containing unsafe
references. It does so by insisting that regions $\rbar$ satisfy the
constraints ($\phi$) imposed by the class on its region parameters.
The later is enforced by checking the validity of $\phi$, with actual
region arguments substituted for formal region parameters, under the
conditions ($\phicx$) guaranteed by the context. The semantics of this
sequent is straightforward, and follows directly from the properties
of outlives and equality relations. For any well-formed core type $T$,
$T@\rgn$ is a well-formed region type if $\rgn$ is a valid region.
The type $\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
$\toprgn$ is a special immortal region that outlives every other live
region. This arrangement allows \C{\RgnZ} handlers to be aliased and
referenced freely from objects in various regions, regardless of their
lifetimes. On the flip side, this also opens up the possibility of
references between transferable regions, which become unsafe in
context of the recipient's address space. Fortunately, such references
are explicitly prohibited by the type rule of \C{\RgnZ} objects, as
described below.

@@

The type rule for \C{new \RgnZ} expressions expects the \RgnZ class's
constructor to be called with a nullary function that returns a value
in its allocation context.  This ensures that the value returned by
the function stores no references to objects allocated elsewhere,
including the top region ($\toprgn$), thus preventing cross-region
references originating from transferable regions. The body of the
function might however create new regions while execution, but this is
not a problem as long as such regions, and objects allocated in them,
don't find their way into the result of its evaluation.

@@

The type rule for \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\pi \notin \Delta$). This condition is needed in order
to prevent the new region from incorrectly assuming existing outlives
relationships on an eponymous region. The expression ($e$) under
\C{letregion} is typechecked with the new region as its allocation
context, under the assumption that the region is live ($\Delta \cup
\{\pi\}$) and that it is outlived by all existing live regions
($\Delta \outlives \pi$). The result of a \C{letregion} expression
must have a type that is well-formed under a context not containing
the new region. This ensures that the value obtained by evaluating a
\C{letregion} expression contains no references to the temporary
objects inside the region.

The rule for \C{open} expression, unlike the rule for \C{letregion},
does not introduce any outlives relationship between the newly opened
region and any pre-existing region while checking the type of the
expression ($e$) under \C{open}. This prevents new objects allocated
inside the transferable region from storing references to those
outside. The newly opened region becomes the allocation context for
$e$, which is type checked under an environment ($\env$) extended with
the type binding for the root object.

The type rule for lambda expression typechecks the lambda-bound
expression ($e$) under an extended type environment containing
bindings for function's arguments, assuming that region parameters are
live, and that declared constraints over region parameters hold. The
constraints ($\phi$) are required to be well-formed under $\Delta$
extended with the function's region parameters ($\rho\rhobar$).
Unlike a typical object, a function closure might capture references
that may become unsafe if the closure escapes its allocation context.
\FB prevents this scenario by requiring the function closure to always
be allocated in the current allocation context ($\rgn$).

@@

The method well-formedness rule makes use of an auxiliary definition
$\override$ that judges whether the current method is a valid
overriding of any eponymous method from the super class. The type
environment is extended with a binding that binds the $\thisZ$ keyword
to the type of the current class. Note that the the type of the
current method as accessed via $\thisZ$ (i.e., $\thisZ.\C{m}$) is
region-parametric, thus admitting region-polymorphic recursion (i.e.,
recursive call to $\C{m}$ can have different region arguments than
$\C{m}$).

@@

The operational semantics of \FB describes a non-trivial runtime
component that introduces memory regions and performs run-time
verification of their typestate. Fig.~\ref{fig:syntax-extra} shows the
additional language constructs of \FB that manifest only at run-time.
A location ($\loc$) abstracts all the memory locations associated with
a region (i.e., each memory region is associated with a single
location). A transferable region can be in one of three possible
states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
manifestations of \C{let} and \C{open}; \C{let} reduces to \C{letd}
while allocating a (static) region, and \C{open} reduces to \C{opened}
while opening a (transferable) region. An \C{opened}
expression is tagged with the typestate ($s$) of the transferable
region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
transferable region can be \C{open}'d from either an open state or a
closed state). Special value $\bot$ denotes an exception.

@@

{\bf A note on region handlers}. \FB stores the region handlers in a
distinguished top region, which abstracts the GC heap. Type system
prevents transferable regions from storing references into the GC
heap. A static region however can store references to region handles,
which need to be taken into account while GC'ing the heap.  Traversing
the region memory to identify references into the GC heap
unfortunately defeats the purpose of regions. A solution is to disable
collecting region handles as long as any static region is open. Since
static regions are intended to store temporary objects while
populating a dynamic region, their lifetimes are short enough to let
region handles to be GC’d.

>>>>

letregion $\pi_0$ in
  let snd = new Object<$\pi_0$>() in
  letregion $\pi_1$ in
    let fst = new Object<$\pi_1$>() in
    let p = new Pair<Object,Object><$\pi_1$,$\pi_1$,$\pi_0$> (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rho_0$ and $\rho_1$ with
$\pi_1$, and $\rho_2$ with $\pi_0$ is allowed because (a) $\pi_0$
and $\pi_1$ are live during the instantiation, and (b). $\pi_0
\outlives \pi_1$ and $\pi_1 \outlives \pi_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\pi_1$, and (b). it
holds references to objects allocated in region $\pi_0$ and $\pi_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\pi_1$, then \C{p} would be contained in $\pi_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\pi_1$,$\pi_1$,$\pi_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\pi_1$}. In general, we treat
$B\inang{\tbar}@\pi$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\pi$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\pi$.

Classes in \FB are independently parameterized over types and regions.
While this design decision has a downside that it allows type
variables to only denote the types of objects contained in a single
region, it yields benefits that outweigh the costs. In particular, it
lets us support region-polymorphic higher-order functions as class
fields.  This allows, for example, a generic class, whose type
parameters are $a$ and $b$, to contain a region-polymorphic function
of type $\inang{\rho_0,\rho_1,\rho_2 \,|\, \rho_1 \outlives \rho_0
\conj \rho_2 \outlives \rho_0}(a@\rho_1,b@\rho_2) \rightarrow
\C{Pair}\inang{a,b}\inang{\rho_0,\rho_1,\rho_2}$ as its field. Such
region-polymorphic higher-order fields are used frequently by the
dataflow operators, which apply them in the context of various regions
(\emph{e.g.,} see Fig.~\ref{fig:motivating-eg-in-broom}). Keeping type
and region parameterizations separate also simplifies the type system
so that inference becomes practical (Sec.~\ref{sec:type-inference}).
The need for polymorphism at the field level is also why \FB treats
function closures specially, rather than as objects of type \C{Func}
as in C\#.

@@

Well-formedness and typing rules of \fbname establish the conditions
under which a region type is considered well-formed, and an expression
is considered to have a certain region type, respectively.
Fig.~\ref{fig:fb-staticsem} contains an illustrative subset of such
rules\footnote{Full formal development can be found in the
appendix}. The rules refer to a context ($\A$), which is a tuple of:
\begin{itemize}
\item A set ($\rhoenv \in 2^\rgn$) of regions that are estimated to be
live,
\item A finite map ($\aenv \in \tyvar \mapsto \fgjN$) of type
variables to their \emph{bounds}, i.e., classes they are declared to
extend (this artifact is inherited from FGJ), and
\item A constraint formula ($\phicx$) that captures the outlives
constraints on regions in $\rhoenv$.
\end{itemize}
In addition, the context for the expression typing judgment also
includes (a).~a type environment ($\env \in x \mapsto \tau$) that
contains the type bindings for variables in scope, and (b).~the region
($\rgn$) that serves as the allocation context for the expression
being type checked. Like the judgments in FGJ~\cite{fgj}, all the
judgments defined by the rules in Fig.~\ref{fig:fb-staticsem} are
implicitly parameterized on a class table ($CT \in cn \mapsto D$) that
maps class names to their definitions in \FB.

@@

\noindent The well-formedness judgment on region types
($\tywf{\A}{\tau}$) makes use of the well-formedness and subtyping
judgments on core types. We use a double-piped turnstile ($\Vdash$)
for judgments in FGJ~\cite{fgj}, and a simple turnstile ($\vdash$) for
those in \FB. The class table ($\absof{CT}$) for FGJ judgments is
derived from \FB's class table ($CT$) by erasing all region
annotations on types, and region arguments in expressions
($\absof{\cdot}$ denotes the region erasure operation). 
% For judgments in FGJ, we use a
% class table that maps class names to their region-erased definitions.
% We denote the region-erased class table as ($\absof{CT}$). The precise
% semantics of region erasure ($\absof{\cdot}$) are defined
% in~\cite{techrep}, but it can be understood as an operation that
% erases region annotations on types. The class table is assumed to bind
% the name $\RgnZ$ to the stub shown in Fig.~\ref{fig:region-stub}.
% This mostly eliminates the need to treat $\RgnZ$ objects, which double
% up as transferable region handlers, as a special case in semantics.
The well-formedness rule for class types
($B\inang{\tbar}\inang{\rbar}$) is responsible for enforcing the
safety property that prevents objects from containing unsafe
references. It does so by insisting that regions $\rbar$ satisfy the
constraints ($\phi$) imposed by the class on its region parameters.
The latter is enforced by checking the validity of $\phi$, with actual
region arguments substituted for formal region parameters, under the
conditions ($\phicx$) guaranteed by the context. The semantics of this
sequent is straightforward, and follows directly from the properties
of outlives and equality relations. For any well-formed core type $T$,
$T@\rgn$ is a well-formed region type if $\rgn$ is a valid region.
The type $\RgnZT{\rgn}$ is well-formed only if $\rgn = \toprgn$, where
$\toprgn$ is a special immortal region that outlives every other live
region. This arrangement allows \C{\RgnZ} handlers to be aliased and
referenced freely from objects in various regions, regardless of their
lifetimes. On the flip side, this also opens up the possibility of
references between transferable regions, which become unsafe in
context of the recipient's address space. Fortunately, such references
are explicitly prohibited by the type rule of \C{\RgnZ} objects, as
described below.

@@

The type rule for the \C{new \RgnZ} expression expects the \RgnZ class's
constructor to be called with a nullary function that returns a value
in its allocation context.  This ensures that the value returned by
the function stores no references to objects allocated elsewhere,
including the top region ($\toprgn$), thus preventing cross-region
references originating from transferable regions. The body of the
function might however create new regions while execution, but this is
not a problem as long as such regions, and objects allocated in them,
don't find their way into the result of its evaluation.

@@

The type rule for the \C{letregion} expression requires that the static
identifier introduced by the expression be unique under the current
context (i.e., $\pi \notin \Delta$). This condition is needed in order
to prevent the new region from incorrectly assuming existing outlives
relationships on an eponymous region. The expression ($e$) under
\C{letregion} is typechecked with the new region as its allocation
context, under the assumption that the region is live ($\Delta \cup
\{\pi\}$) and that it is outlived by all existing live regions
($\Delta \outlives \pi$). The result of a \C{letregion} expression
must have a type that is well-formed under a context not containing
the new region. This ensures that the value obtained by evaluating a
\C{letregion} expression contains no references to the temporary
objects inside the region.

The rule for the \C{open} expression, unlike the rule for
\C{letregion}, does not introduce any outlives relationship between
the newly opened region and any pre-existing region while checking the
type of the expression ($e$) under \C{open}. This prevents new objects
allocated inside the transferable region from storing references to
those outside. The newly opened region becomes the allocation context
for $e$, which is type checked under an environment ($\env$) extended
with the type binding for the root object.

The type rule for the lambda expression typechecks the lambda-bound
expression ($e$) under an extended type environment containing
bindings for function's arguments, assuming that region parameters are
live, and that declared constraints over region parameters hold. The
constraints ($\phi$) are required to be well-formed under $\Delta$
extended with the function's region parameters ($\rho\rhobar$).
Unlike a typical object, a function closure might capture references
that may become unsafe if the closure escapes its allocation context.
\FB prevents this scenario by requiring the function closure to always
be allocated in the current allocation context ($\rgn$).

@@

The method well-formedness rule makes use of an auxiliary definition
$\override$ that judges whether the current method is a valid
overriding of any eponymous method from the super class. The type
environment is extended with a binding that binds the $\thisZ$ keyword
to the type of the current class. Note that the type of the
current method as accessed via $\thisZ$ (i.e., $\thisZ.\C{m}$) is
region-parametric, thus admitting region-polymorphic recursion (i.e.,
recursive call to $\C{m}$ can have different region arguments than
$\C{m}$).

@@

The operational semantics of \FB describes a non-trivial runtime
component that introduces memory regions and performs run-time
verification of their typestate. Fig.~\ref{fig:fb-syntax-extra} shows the
additional language constructs of \FB that manifest only at run-time.
A location ($\loc$) abstracts all the memory locations associated with
a region (i.e., each memory region is associated with a single
location). A transferable region can be in one of three possible
states: closed ($\USED$), open/live ($\LIVE$), and transferred/freed
($\XFERRED$). Constructs \C{letd} and \C{opened} are the run-time
manifestations of \C{let} and \C{open}; \C{let} reduces to \C{letd}
while allocating a (static) region, and \C{open} reduces to \C{opened}
while opening a (transferable) region. An \C{opened}
expression is tagged with the typestate ($s$) of the transferable
region before it is opened (as Fig.~\ref{fig:region-fsm} shows, a
transferable region can be \C{open}'d from either an open state or a
closed state). Special value $\bot$ denotes an exception.

@@

{\bf Integrating regions with GC heap}. \FB stores the region handlers in a
distinguished top region, which abstracts the GC heap. Type system
prevents transferable regions from storing references into the GC
heap. A static region however can store references to region handles,
which need to be taken into account while GC'ing the heap. Traversing
the region memory to identify references into the GC heap
unfortunately defeats the purpose of regions. The solution we adopt is
to disable collecting region handles as long as any static region is
open. Since static regions are intended to store temporary objects
while populating a dynamic region, their lifetimes are short enough to
let region handles to be GC’d.
